Article Title

### 2014-05-07 plos.tex
added section on variance estimation

We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best represented as the combination of a sparse matrix of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate expressed a graph of interactions organized with respect to the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas the negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this ``sparse + latent'' estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical representations of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by green spheres and latent units by light-shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} with several latent units. \\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with two latent units, {\bf C.}  a correlation matrix with 85\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-1 (\ie one latent unit) matrix and a sparse matrix with 89\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=1000$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. Some exceptions occur when the sample is too small to reveal the structure in the data. In these cases estimators with simpler structures can be more efficient than estimators with true structure.
\\
{\bf Row 6:} Relative cross-validation losses (Eq.~\ref{eq:rel-cv-loss}) for the five estimators with respect to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:1}
\end{FPfigure}

@@

The target of estimator $C_{\sf factor}$ is the factor model $L + D$, where $L$ is the $p\times p$ positive \hl{semidefinite} matrix with low rank and $D$ is the diagonal matrix of independent variances.  The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = (1-\lambda) C_{\sf sample} + \lambda (L + D),
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{E}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with between 100 and 300 repetitions of each direction. In this stimulus, each grating lasted 1 second and was followed by a 1-second blank.  The average stimulus response was subtracted from each trial. The traces were then binned into 150 ms intervals for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low (average across sites of 0.02 and \hl{less than 0.04 in any site}) (\figref{2}{G} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--G.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals from eight cells out of 298 cells downsampled to 20 Hz. The inferred firing rate binned in 150 ms intervals are indicated by red ticks below each trace.
    {\bf D.} The raster plot of the inferred firing rates, binned in 150 ms intervals, from the entire population from the first (left) and last (right) minute of the entire recording.  The traces from {\bf C} are highlighted in red.
    {\bf E.} The spatial arrangement and orientation tuning of the 298 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf F.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf G.} Histogram of noise correlation coefficients. The red line indicates the mean.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimators that explicitly modeled the partial correlations between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs, to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

@@

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
    {\bf Cross-validation losses of estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ relative to $C_{\sf sparse+latent}$.}
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<10^{-5}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

Furthermore, $C_{\sf sparse}$ outperformed all the other estimators besides $C_{\sf sparse+latent}$ (Fig.~S\ref{supp:01}) suggesting the greater importance of modeling interactions between pairs of neurons than the common fluctuations of activity across the entire population.

The same evaluation based on the quadratic loss function,
\begin{equation}\label{eq:quadratic}
\loss{C,C_{\sf sample}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf sample}^\prime-I)^2,
\end{equation}
instead of the normal loss function reproduced the same relationship between the estimators (Fig.~S\ref{supp:02}). This suggests that the results of this study are robust to the choice of the loss function and do not depend on the assumption of gaussianity.

@@

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure06.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations and regularized partial correlations on orientation tuning and physical distance.}
    {\bf A--C.} Average partial correlations (red) estimated by $C_{\sf sparse+latent}$ and average sample correlations (black) averaged across multiple imaged sites. At each site the correlations were normalized by the respective average correlation shown in \figref{5}{D}.  The number of sites, $n$, that qualified to be included in the analysis is indicated in each panel. Sites were included if they had at least 20 pairs of neurons in each of the intervals. The error bars indicate the standard error of the mean based on on the number of sites $n$.
    {\bf A.} Average correlations between pairs of neurons tuned to orientation with differences in preferred orientation in the intervals of 0--15$^\circ$, 15--45$^\circ$ and 45--90$^\circ$. 
    {\bf B.} Average correlations between pairs of neurons located at the same depth ($\pm$25$\mu$m) separated by lateral distances in the intervals of 0--25 $\mu$m, 25--75 $\mu$m, 75--150 $\mu$m, and 150+ $\mu$m.
    {\bf C.} Average correlations between pairs of neurons displaced laterally by less than 25 $\mu$m separated in depth by distances in the intervals of 0--25 $\mu$m, 25--60 $\mu$m, and 60+ $\mu$m.
    {\bf D--F.} Normalized connectivity of positive (green) and negative (dark red) interactions from the sparse component obtained from $C_{\sf sparse+latent}$. Normalized connectivity was computed as the fraction of pairs connected by interactions of corresponding signs in each interval divided by the fraction of non-zero interactions across the entire site. }
\label{fig:6}
\end{figure}

@@

The collected fluorescent traces were deconvolved to reconstruct the firing rates for each neuron. First, the first principal component was subtracted from the traces, which reduced common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were low-pass filtered below 0.1 Hz and downsampled to 20 Hz. Firing rates were estimated using a fast non-negative deconvolution algorithm \cite{Vogelstein:2010}.

@@

For covariance estimation, the analysis was limited to the period with 2 or 5 stimulus conditions and lasted between 14 and 27 minutes (mean 22 minutes).  Cells that did not have substantial spiking activity (those whose variance was less than 2\% of the $90^{th}$ percentile in each site) were excluded from the analysis.

@@

Estimator $C_{\sf factor}$ used two hyperparameters: the number of latent factors $d$ and the shrinkage intensity $\lambda \in [0, 1]$.  
The matrix $L$ of rank $d$ and the diagonal matrix of individual variances $D$ were computed by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\tilde L,\tilde D} \loss{\tilde L + \tilde D,C_{\sf sample}},
\end{equation}
which we solved by an expectation-maximization (EM) algorithm.  Under our chosen loss function (Eq.~\ref{eq:loss}), this is equivalent to maximum likelihood estimation of $L$ and $D$ under the multivariate Gaussian distribution.  The final regularized estimate is obtained by linear shrinkage toward the factor model (Eq.~\ref{eq:c-factor}).

Estimator $C_{\sf sparse}$  has one hyperparameter $\lambda$ to regulate the sparsity of its precision matrix $S$. The precision matrix is computed in two steps: First, the zero structure is determined by minimizing the $L_1$-penalized loss:
\begin{equation}
Z = \argmin\limits_{\tilde Z \succ 0} \loss{{\tilde Z}^{-1},C_{\sf sample}} + \lambda \|\tilde Z \|_1
\end{equation}
where $\tilde Z\succ 0$ denotes the constraint that $\tilde Z$ be a positive definite matrix and $\|\tilde Z\|_1$ is the element-wise $L_1$ norm of the matrix $\tilde Z$. This problem formulation is known as \emph{graphical lasso} \cite{Meinshausen:2006, Friedman:2008}. To solve this minimization problem, we modified the alternative-direction method of multipliers (ADMM) algorithm \cite{Ma:2013}. 
Then, after the zero structure was determined, the remaining coefficients were fitted without penalty:
\begin{equation}
S = \argmin\limits_{\tilde S \in Z^\sharp} \loss{\tilde S,C_{\sf sample}},
\end{equation}
where $Z^\sharp$ denotes the set of positive-definite matrices with the same zero structure as $Z$.  This step was also solved by ADMM.  Then the final estimate is the inverse of $S$ (Eq.~\ref{eq:c-sparse}). Unlike $C_{\sf diag}$ and $C_{\sf factor}$, this estimator does not include linear shrinkage: the selection of the sparsity level provides sufficient flexibility to fine tune the regularization level.

Estimator $C_{\sf sparse+latent}$ has two hyperparameters: the number of latent units $d$ and \hl{hyperparameter $\lambda$ regulating the sparsity level}. $C_{\sf sparse+latent}$ estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units and expresses their partial correlation structure, and $S_{12}$ and $S_{22}$ are of size $p\times d$ and $d\times d$, respectively. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ has rank $d$. Rather than searching for the optimal sparse structure of $S_{12}$ and $S_{22}$, an ill-posed problem, we estimated these components together as the low-rank matrix $L$.

The estimate is found in two steps. First, we use the ADMM algorithm to find the zero structure of $S$ by minimizing the $L_1$-penalized loss \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
(Z,\cdot) = \argmin\limits_{\tilde Z,\tilde L} \loss{\tilde Z-\tilde L, C_{\sf sample}} + \lambda\|\tilde Z\|_1
\end{equation}
Then we find the sparse and low-rank components of the precision matrix by solving (also using ADMM) 
\begin{equation}
(S,L) = \argmin\limits_{\tilde S \in Z^\sharp,\tilde L} \loss{\tilde S-\tilde L, C_{\sf sample} }
\end{equation}

@@

%\section*{Figure Legends}
\newpage
\section*{Supporting Information}
\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/src/Supp1.eps}
\end{center}
\caption{Comparison of the $C_{\sf sparse}$ estimator to the other estimators with respect to the normal loss function.}
\label{supp:01}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/src/Supp2.eps}
\end{center}
\caption{Comparison of the $C_{\sf sparse+latent}$ estimator to the other estimators with respect to the quadratic loss.}
\label{supp:02}
\end{figure}

>>>>

We sought statistically efficient estimators of neural correlation matrices in recordings from large, dense groups of cortical neurons.  Using fast 3D random-access laser scanning microscopy of calcium signals, we recorded the activity of nearly every neuron in volumes 200 $\mu$m wide and 100 $\mu$m deep (150--350 cells) in mouse visual cortex.  We hypothesized that in these dense recordings, the correlation matrix should be best represented as the combination of a sparse matrix of pairwise partial correlations representing local interactions and a low-rank component representing common fluctuations and external inputs.  Indeed, in cross-validation tests, the covariance matrix estimator with this structure consistently outperformed other regularized estimators. The sparse component of the estimate defined a graph of interactions. These interactions reflected the physical distances and orientation tuning properties of cells: The density of positive \sq{excitatory} interactions decreased rapidly with geometric distances and with differences in orientation preference whereas negative \sq{inhibitory} interactions were less selective.  Because of its superior performance, this \sq{sparse + latent} estimator likely provides a more physiologically relevant representation of the functional connectivity in dense recordings than the sample correlation matrix.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure01.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
    {\bf Row 1.} Graphical representations of the target estimates of the four respective regularized covariance matrix estimators.  Recorded neurons are represented by green spheres and latent units by light-shaded spheres.  Edges represent non-zero partial correlations, \ie \sq{interactions}.
        {\bf Row 1, A}.  For estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear dependencies. 
        {\bf  Row 1, B.} For estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), representing systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. For estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from a sparse set of  linear associations between the observed units. 
        {\bf  Row 1, D}.  For estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to sparse associations between the recorded cells \emph{and} with several latent units. \\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with two latent units, {\bf C.}  a correlation matrix with 85\%  off-diagonal zeros in its inverse, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-1 (\ie one latent unit) matrix and a sparse matrix with 89\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=1000$ drawn from simulated random processes with respective correlation matrices shown in Row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) for the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with structure that matches the true model converged to zero faster than the other estimators. Some exceptions occur when the sample is too small to reveal the structure in the data. In these cases estimators with simpler structures can be more efficient than estimators with true structure.
\\
{\bf Row 6:} Cross-validation losses for the five estimators relative to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:1}
\end{FPfigure}

@@

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive \hl{semidefinite} matrix with low rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned to the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0066 to 0.043, mean 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals from eight cells out of 298 cells downsampled to 20 Hz. Spiking activity inferred by non-negative deconvolution is shown below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 298 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.017.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimators that explicitly modeled the partial correlations between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

@@

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
   {\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to the normal loss function (eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<10^{-4}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

Furthermore, estimator $C_{\sf sparse}$ outperformed estimator $C_{\sf factor}$ (Fig.~S\ref{supp:01}) suggesting the greater importance of modeling interactions between pairs of neurons than the common fluctuations of activity across the entire population.

The same evaluation based on the quadratic loss function,
\begin{equation}\label{eq:quad}
\loss{C,C_{\sf sample}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf sample}^\prime-I)^2,
\end{equation}
instead of the normal loss function reproduced the same relationship between the estimators (Fig.~S\ref{supp:02}). This suggests that the results of this study are robust to the choice of the loss function and do not depend on the assumption of gaussianity.

@@

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure06.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations and regularized partial correlations on orientation tuning and physical distance.}
    {\bf A--C.} Average partial correlations (red) estimated by $C_{\sf sparse+latent}$ and average sample correlations (black) averaged across multiple imaged sites. At each site the correlations were normalized by the respective average correlation shown in \figref{5}{D}.  The number of sites, $n$, that qualified to be included in the analysis is indicated in each panel. Sites were included if they had at least 20 pairs of neurons in each of the intervals. The error bars indicate the standard error of the mean based on on the number of sites $n$.
    {\bf A.} Average correlations between pairs of neurons tuned to orientation with differences in preferred orientation in the intervals of 0--15$^\circ$, 15--45$^\circ$ and 45--90$^\circ$. 
    {\bf B.} Average correlations between pairs of neurons located at the same depth ($\pm$30$\mu$m) separated by lateral distances in the intervals of 0--25 $\mu$m, 25--75 $\mu$m, 75--150 $\mu$m, and 150+ $\mu$m.
    {\bf C.} Average correlations between pairs of neurons displaced laterally by less than 30 $\mu$m separated in depth by distances in the intervals of 0--25 $\mu$m, 25--60 $\mu$m, and 60+ $\mu$m.
    {\bf D--F.} Normalized connectivity of positive (green) and negative (dark red) interactions from the sparse component obtained from $C_{\sf sparse+latent}$. Normalized connectivity was computed as the fraction of pairs connected by interactions of corresponding signs in each interval divided by the fraction of non-zero interactions across the entire site. }
\label{fig:6}
\end{figure}

@@

The collected fluorescent traces were deconvolved to reconstruct the firing rates for each neuron. First, the first principal component was subtracted from the traces, which reduced common mode noise related to small cardiovascular movements \cite{Cotton:2013}. The resulting traces were low-pass filtered below 0.1 Hz and downsampled to 20 Hz (\figref{2}{C}). Firing rates were estimated using a fast non-negative deconvolution algorithm \cite{Vogelstein:2010}.

@@

For covariance estimation, the analysis was limited to the period with 2 or 5 stimulus conditions and lasted between 14 and 27 minutes (mean 22 minutes).  Cells that did not have substantial spiking activity (those whose variance was less than 1\% of the median across the site) or whose activity was unstable (those whose variance in the least active quarter of the recording did not exceed 1\% of the variance in the most active quarter) were excluded from the analysis.

@@

Estimator $C_{\sf factor}$ used two hyperparameters: the number of latent factors $d$ and the shrinkage intensity $\lambda \in [0, 1]$.  
The matrix $L$ of rank $d$ and the diagonal matrix of individual variances $D$ were computed by solving the minimization problem
\begin{equation}
(L,D) = \argmin\limits_{\hat L,\hat D} \loss{\hat L + \hat D,C_{\sf sample}},
\end{equation}
which we solved by an expectation-maximization (EM) algorithm.  Under our chosen loss function (Eq.~\ref{eq:loss}), this is equivalent to maximum likelihood estimation of $L$ and $D$ under the multivariate Gaussian distribution.  The final regularized estimate is obtained by linear shrinkage toward the factor model (Eq.~\ref{eq:c-factor}).

Estimator $C_{\sf sparse}$  has one hyperparameter $\lambda$ to regulate the sparsity of its precision matrix $S$. The precision matrix is obtained by minimizing the $L_1$-penalized loss:
\begin{equation}
S = \argmin\limits_{\hat S \succ 0} \loss{{\hat S}^{-1},C_{\sf sample}} + \lambda \|\hat S \|_1
\end{equation}
where $\hat S\succ 0$ denotes the constraint that $\hat S$ be a positive definite matrix and $\|\hat S\|_1$ is the element-wise $L_1$ norm of the matrix $\hat S$. This problem formulation is known as \emph{graphical lasso} \cite{Meinshausen:2006, Friedman:2008}. To solve this minimization problem, we adapted the alternative-direction method of multipliers (ADMM) algorithm \cite{Ma:2013}. 
Then the final estimate is the inverse of $S$ (Eq.~\ref{eq:c-sparse}). Unlike $C_{\sf diag}$ and $C_{\sf factor}$, this estimator does not include linear shrinkage: the selection of the sparsity level provides sufficient flexibility to fine tune the regularization level.

Estimator $C_{\sf sparse+latent}$ has two hyperparameters: the number of latent units $d$ and \hl{hyperparameter $\lambda$ regulating the sparsity level}. $C_{\sf sparse+latent}$ estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.  
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units and expresses their partial correlation structure, and $S_{12}$ and $S_{22}$ are of size $p\times d$ and $d\times d$, respectively. 
Then the covariance matrix of the observed population is 
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ has rank $d$. Rather than searching for the optimal sparse structure of $S_{12}$ and $S_{22}$, an ill-posed problem, we estimated these components together as the low-rank matrix $L$. We solved this problem by adapting the ADMM algorithm to minimizing the loss function with penalty on the $L_1$-norm  of $S$ and the nuclear norm of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta \Tr{\hat L}
\end{equation}

@@

%\section*{Figure Legends}
\newpage
\section*{Supporting Information}
\subsection*{Appendix: Estimation of variance}
One decision that was made in this paper was to model a constant covariance matrix across different conditions. In other words, the variances was modeled as a constant across conditions. In contrast, it is more common to estimate correlations separately in different conditions and average the correlations coefficients from different conditions. In this appendix we justify this decision.

Let's consider two random variables $x_1$ and $x_2$ with zero means and respective variances $v_1^2$ and $v_2^2$ in one condition and $a^2v_1^2$ and $b^2v_2^2$ in the second condition. Assume that the correlation coefficient $\rho$ is the same in both conditions.

Let $\langle \cdot \rangle_i$ denote the expectation in condition $i$. Then
\begin{align*}
\langle x_1^2 \rangle_1 = v_1^2 && \langle x_2^2 \rangle_1 = v_2^2 && \rho = \frac{\langle x_1 x_2 \rangle_1}{v_1v_2}\\ 
\langle x_1^2 \rangle_2 = a^2v_1^2 && \langle x_2^2 \rangle_2 = b^2v_2^2  && \rho = \frac{\langle x_1 x_2 \rangle_2}{ab\,v_1v_2} 
\end{align*}

If we ignore the differences between the variances in the two conditions, assuming equal numbers of trials in each condition, the respective  variances for the two variables will be 
\begin{align*}
\tilde v_1^2 = \frac 1 2 \left( v_1^2 + a^2v_1^2 \right) = \frac{1+a^2}{2} v_1^2
&&
\tilde v_2^2 = \frac 1 2 \left( v_2^2 + b^2v_2^2 \right) = \frac{1+b^2}{2} v_2^2
\end{align*}
The resulting correlation coefficient will be
\begin{equation*}
\tilde \rho = \frac{\langle x_1 x_2 \rangle_1 + \langle x_1 x_2 \rangle_2} {2 \tilde v_1 \tilde v_2}
= \rho \frac{1 + ab}{\sqrt{1+a^2}\sqrt{1+b^2}}
= \rho \sqrt{\frac{1+2ab+a^2b^2}{1+a^2+b^2+a^2b^2} }
\end{equation*}
Note that $\tilde \rho = \rho$ iff $a=b$ and $\tilde \rho<\rho$ when $a\ne b$.  In other words, when stimulus conditions modulate the variances by different amounts, the assumption of constant variance leads to the underestimation of the correlation between differently tuned pairs of cells. This effect may lead to the overestimation of the effect of common tuning on correlations (`limited-range correlations').

\subsection*{Supplementary Figures}
\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/src/Supp1.eps}
\end{center}
\caption{
{\bf Performance of estimator $C_{\sf sparse}$ with respect to normal loss (Eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$}. 
Estimator $C_{\sf sparse}$ outperformed $C_{\sf sample}$ ($p=5\cdot 10^{-6}$) and $C_{\sf factor}$ ($p=.01$), but not $C_{\sf diag}$ ($p=.17$).}
\label{supp:01}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics{./figures/src/Supp2.eps}
\end{center}
\caption{
{\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to \emph{quadratic loss} (Eq.~\ref{eq:quad}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
All comparisons showed significant advantage of estimator $C_{\sf sparse+latent}$ ($p<10^{-3}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice).}
\label{supp:02}
\end{figure}
